Discrete Ordinates

### 2017-06-03 docprep/chapmeth.tex
Finished draft of the thesis through the results
Minor fixes and added some figs

%%
%% This is file `chapmin.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% ths.dtx  (with options: `chapmin')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from chapmin.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file ths.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)

@@

The linear Boltzmann equation (LBE) is generally true of any particle given that outside forces (electromagnetic or gravitational) are either not present or negligible and particle-particle collisions are insignificant. Therefore, the LBE is generally not valid for charged particles or systems whose particle density is on the order of the medium they are traveling through. In such systems, particle-particle interactions may not be negligible.

@@

The steady state form of the LBE can be readily derived from simple intuition. In the steady state, particle production and removal must be equal. In any volume, three production mechanisms are present. Particles can (1) stream into the volume across a bounding surface, (2) inscatter from another energy-direction component of phase space, or (3) be produced directly by the external source. Three removal terms are also present, particles can (1) stream out of the volume of interest by crossing a bounding surface, (2) be absorbed and vanish entirely, or (3) outscatter to a energy-direction component of phase space other than the one of interest.

The downscatter and absorption removal terms are typically grouped together using
\begin{equation}
\Sigma_t = \Sigma_a + \Sigma_s
\end{equation}
where $\Sigma_t$, $\Sigma_a$, and $\Sigma_s$ are the macroscopic total, absorption, and scatter cross sections respectively. The removal of particles from interactions (scatter and absorption) per volume is then
\begin{equation}
\Sigma_t(\boldsymbol{r}, E) \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}
where $\psi$ is the angular flux. The production and removal streaming operators are combined as 
\begin{equation}
\hat{\Omega} \cdot \nabla \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}
and included as a removal term since the normal is defined pointing outward from the volume. The normal sign convention results in particles streaming out being positive and those streaming in becoming negative.

@@

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/coord_sys}
  \end{center}
  \caption{The coordinate system used in DOCTORS. Given an arbitrary direction, $\Omega$, $\mu$, $\eta$, and $\xi$ are its direction cosines with respect to the $x$, $y$, and $z$ axes respectively. $\varphi$ is the azimuthal angle (with respect to $x$ and the d $\theta$ is the polar angle (with respect to $z$).}
\label{fig:coord_sys}
\end{figure}

The rotation symmetrical quadrature is implemented in DOCTORS. From amongst many different quadrature sets, the rotation symmetrical quadratures were selected because they are easy to implement, rotationally symmetric, and the most commonly used in production discrete ordinate solvers [CITE] which enables simple, direct comparison to other solvers. Arbitrary quadratures are permissible in a plain text file supplied by the user enabling other quadratures.

THEORY AND MATH?

To include a user defined quadrature, the plain text file should be formatted such that it contains four columns similar to Table~\ref{tab:quad_format} (without any headings). No checks are performed to ensure that octants are covered, this allows simple testing quadratures that utilize only a single direction to be used. Values for $\mu$, $\eta$, and $\xi$ do not necessarily need to add to unity as they will be automatically normalized by DOCTORS. Likewise, the weights will be normalized such that their summation is unity. However, user defined directions \textit{should not} be parallel to any major axis. Doing so may result in undefined behavior. Also, after normalization, no two rows should have identical values for $\mu$, $\eta$, and $\xi$ such as the last two rows of Table~\ref{tab:quad_interp}, nor may any weights be negative. The data in Table~\ref{tab:quad_format} would be read in and renormalized by DOCTORS. The normalized data would be identical to the data in Table~\ref{tab:quad_interp}.

Values are read in as a 32 bit IEEE-754 floating point number [CITE] which ensures seven significant digits. Significant digits beyond this will be truncated. Note that Table~\ref{tab:quad_interp} rounds four digits after the decimal.

@@

A given direction, $\hat{\Omega}$, is determined by its three cosine components: 
\begin{equation} \label{eq:omega_cos}
\hat{\Omega} = \mu \hat{i} + \eta \hat{j} + \xi \hat{k}
\end{equation}
where $\hat{i}$, $\hat{j}$, and $\hat{k}$ are the unit directions in the $x$, $y$, and $z$ directions respectively. Therefore, any discrete direction $\hat{\Omega}_a$ can be expressed as $<\mu, \eta, \xi>$. The two forms will be used interchangeably.

Discretizing Eq.~\ref{eq:boltz} in angular space gives
\begin{equation} \label{eq:boltz_a}
\begin{split}
&\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t(\boldsymbol{r}, E) \right]
\psi_{a}(\boldsymbol{r}, E) = \\
&\sum_{a=0}^{N_a-1} \int_0^\infty \Sigma_{s, a, a'}(\boldsymbol{r}, E' \rightarrow E) \psi_{a'}(\boldsymbol{r}, E') dE' \omega_a + S_a(\boldsymbol{r}, E)
\end{split}
\end{equation}
where the subscript $a$ denotes the direction number and $\omega_a$ is a weight associated with that direction. The weights are designed with the angles such that once a discrete quadrature is selected, integration over continuous space becomes integration in discrete space approximated by
\begin{equation} \label{eq:disc_int}
\int_{4 \pi} f(\hat{\Omega}) d\hat{\Omega} \approx \sum_{a=0}^{N_a-1} f(\hat{\Omega}_a) \omega_a = 1.
\end{equation}

@@

Once the energy group structure is selected, cross section data must be available. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight some energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries us a uniform function [CITE].
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

In photon based problems, upscatter is typically negligible [CITE]. By the time photons become low enough in energy to upscatter significantly, they are no longer of dosimetric concern. Removing upscatter reduces Eq.~\ref{eq:boltz_e} to
\begin{equation} \label{eq:boltz_e2}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}
which is very similar; only the starting point for the scatter summation changes. Computationally, though, the lack of upscatter makes a significant difference. The highest energy group can be solved independently of all those below it. Once that group is solved, the next group can also be solved and so forth. In the presence of upscatter, the highest group depends on the solution to all groups below it so all energy groups must be solved simultaneously.

@@

The problem domain for a CT voxel phantom is a regular, rectangular parallelepiped of dimension $D_x \times D_y \times D_z$, the values of which must be known based on the CT setup by the user. The mesh is partitioned into $N_x$, $N_y$, and $N_z$ evenly spaced bins along each major direction as shown in Fig.~\ref{fig:spatial_disc}. The total number of voxels, $N_V$ is easily computed from the number of bins in each major direction:
\begin{equation} \label{eq:n_v}
N_V = N_x N_y N_z.
\end{equation}
The length of a voxel along the $x$-direction is computed by 
\begin{equation} \label{eq:mesh_x}
\Delta x = \frac{D_x}{N_x}.
\end{equation}
Analogous equations apply in the $y$ and $z$ directions as well.

@@

Voxels can be indexed in one of two ways, either by their component indices or global index. Each voxel in the mesh has three components indices, one in each major direction, $i_x$, $i_y$, and $i_z$ that range from 0 to $N_x - 1$, $N_y - 1$, or $N_z-1$. However, this makes indexing cumbersome, so they are combined into a unique global index, $i$ using Eq.~\ref{eq:indx_flat}. In general, this is known as "flattening" a matrix into a one-dimensional vector. More details about the flattening implementation used in DOCTORS is given in Section~\ref{sec:flatten}.

\begin{equation} \label{eq:indx_flat}
i = i_x (N_z + N_y) + i_y N_z + i_z
\end{equation}

Using global indexing discussed, the fully discretized the steady state LBE given is
\begin{equation} \label{eq:boltz_i}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_{t,i}^g \right]
\psi_{i,a}^{g} = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g
\end{equation}

@@

To solve the fully discretized form of the Linear Boltzmann Equation given in Eq.~\ref{eq:boltz_i}, the gradient operator must first be computed numerically. Recall that $\Omega_a$ can be written in vector notation using its cosine components as defined in Eq.~\ref{eq:omega_cos}. Also recall the definition of the gradient:
\begin{equation} \label{eq:grad}
\nabla f(x, y, z) = \frac{\partial f(x, y, z)}{\partial x} \hat{i} + \frac{\partial f(x, y, z)}{\partial y} \hat{j} + \frac{\partial f(x, y, z)}{\partial z} \hat{k}.
\end{equation}
To a first order approximation, a partial derivative is computed as
\begin{equation} \label{eq:deriv_1}
\frac{\partial f(\zeta)}{\partial \zeta} \approx \frac{f(\zeta+\Delta \zeta/2) - f(\zeta + \Delta \zeta/2)}{\Delta \zeta}
\end{equation}
with respect to $\zeta$. Combining Eq.~\ref{eq:grad} and~\ref{eq:deriv_1}, the $\hat{\Omega} \cdot \nabla \psi$ term can be rewritten as:
\begin{equation} \label{eq:spatial_1}
\begin{split}
\Omega \cdot \nabla \psi 
& = <\mu, \eta, \xi> \cdot <\frac{\partial \psi}{\partial x}, \frac{\partial \psi}{\partial y}, \frac{\partial \psi}{\partial z}> \\
& =
\frac{\partial \psi}{\partial x}\mu + \frac{\partial \psi}{\partial y}\eta + \frac{\partial \psi}{\partial z}\xi \\
& \approx 
\frac{\psi(x + \Delta x/2, y, z) - \psi(x - \Delta x/2, y, z)}{\Delta x} \mu \\
&+ 
\frac{\psi(x, y + \Delta y/2, z) - \psi(x, y - \Delta y/2, z)}{\Delta y} \eta \\
&+ 
\frac{\psi(x, y, z + \Delta z/2) - \psi(x, y, z - \Delta z/2)}{\Delta z} \xi.
\end{split}
\end{equation}
The flux variables $\psi_{i,a,x,in}^g$, $\psi_{i,a,x,out}^g$, $\psi_{i,a,y,in}^g$, $\psi_{i,a,y,out}^g$, $\psi_{i,a,z,in}^g$, and $\psi_{i,a,z,out}^g$ are \textit{surface} averaged flux values while $\psi_{i,a}^{g}$ is a \textit{volume} averaged flux value.

The gradient is computed with respect to th edirection $\hat{\Omega}$, thus Eq.~\ref{eq:spatial_1} is valid only for the first octant ($\mu, \eta, xi > 0$). In other octants, a new gradient must be used. Rather than enumerating all eight variants, Eq.~\ref{eq:spatial_1} can be generalized to
\begin{equation} \label{eq:deriv_2}
\Omega \cdot \nabla \psi \approx 
\frac{\psi_{x,out} - \psi_{x,in}}{\Delta x} \mu + 
\frac{\psi_{y,out} - \psi_{y,in}}{\Delta y} \eta + 
\frac{\psi_{z,out} - \psi_{z,in}}{\Delta z} \xi
\end{equation}
where
\begin{equation}
\psi_{x,out} = 
\begin{cases}
\psi(x+\Delta x/2, y, z) - \psi(x-\Delta x/2, y, z) \,, \quad \mu > 0 \\
\psi(x-\Delta x/2, y, z) - \psi(x+\Delta x/2, y, z) \,, \quad \mu < 0
\end{cases}
\end{equation}
and
\begin{equation}
\psi_{x,in} = 
\begin{cases}
\psi(x-\Delta x/2, y, z) - \psi(x+\Delta x/2, y, z) \,, \quad \mu > 0 \\
\psi(x+\Delta x/2, y, z) - \psi(x-\Delta x/2, y, z) \,, \quad \mu < 0
\end{cases}
\end{equation}
and $\psi_{y,out}$, $\psi_{y,in}$, $\psi_{z,out}$, and $\psi_{z,in}$ are similarly defined but with respect to $\eta$ and $\xi$ respectively.

@@

The diamond difference approximation, relates the incoming and outgoing surface fluxes to the cell averaged flux. The cell averaged flux is assumed to be the average of any two opposite surface fluxes. This is assumed to be the case in all three spatial dimensions. This leads to:
\begin{equation} \label{eq:dd}
\begin{split}
\frac{\psi_{i,a,x,in}^g + \psi_{i,a,x,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,y,in}^g + \psi_{i,a,y,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,z,in}^g + \psi_{i,a,z,out}^g}{2} &= \psi_{i,a}^{g}.
\end{split}
\end{equation}
Multiplying both sides of Eq.~\ref{eq:dd} by 2 and subtracting twice the incoming surface flux from each yields:
\begin{equation} \label{eq:dd2}
\begin{split}
\psi_{i,a,x,out}^g - \psi_{i,a,x,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,x,in}^g \\
\psi_{i,a,y,out}^g - \psi_{i,a,y,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,y,in}^g \\
\psi_{i,a,z,out}^g - \psi_{i,a,z,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,z,in}^g.
\end{split}
\end{equation}

The form of the left hand side of Eq.~\ref{eq:dd2} is the same as in Eq.~\ref{eq:boltz_i2}. This allows Eq.~\ref{eq:dd2} to be used to make a substitution in Eq.~\ref{eq:boltz_i2} that removes the dependence on the outgoing flux from discretized equation. After the substitution:
\begin{equation} \label{eq:boltz_i3}
\begin{split}
&\left[ 
\frac{2\psi_{i,a}^{g} - 2\psi_{i,a,x,in}^g}{\Delta x} \mu + 
\frac{2\psi_{i,a}^{g} - 2\psi_{i,a,y,in}^g}{\Delta y} \eta + 
\frac{2\psi_{i,a}^{g} - 2\psi_{i,a,z,in}^g}{\Delta z} \xi
\right]
+ \Sigma_{t,i}^g \psi_{i,a}^{g} \\
& = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g.
\end{split}
\end{equation}

@@

\subsection{Convergence Criteria}
The typical convergence criteria used to determine whether or to terminate further iteration is the maximum relative error from the $i-1^{th}$ iteration to the $i^{th}$ iteration given in Eq.~\ref{eq:conv} over the phase space $\mathbb{R}$. Once the relative error in the parameter of interest, $\phi$, is below some threshold, $\epsilon$, the iteration stops and the solution is considered converged.

\begin{equation}\label{eq:conv}
\max_{\mathbb{R}} \left\{ \frac{\phi_i - \phi_{i-1}}{\phi_i} \right\} < \epsilon
\end{equation}

In some problems, the convergence criteria given in Eq.~\ref{eq:conv} is not reached for many iterations. In these cases, terminating the solution early rather than waiting for full convergence is appropriate. This is enforced by a second criterial given by Eq.~\ref{eq:conv2}. Whenever the number of iterations, $i$, exceeds the total permissible number of iterations, $I$, the iteration is terminated.

\begin{equation}\label{eq:conv2}
i > I
\end{equation}

@@

FIGURE TO SHOW CONVERGENCE RATE.

In order to measure the uncertainty due to floating-point error, another convergence criteria was added. Rather than looking at the relative change from iteration to iteration, the total absolute change is measured:
\begin{equation}\label{eq:conv3}
\epsilon = \sum_{\mathbb{R}} |\phi_i - \phi_{i-1}|.
\end{equation}
The termination criteria is given by Eq.~\ref{eq:conv4}. Computation of a value for $\epsilon$ requires two previous iteration, thus this condition cannot be met until the third iteration and thereafter.

\begin{equation}\label{eq:conv4}
\epsilon_i > \epsilon_{i-1}
\end{equation}

@@

The data files containing the scatter cross sections are distributed using a Legendre polynomial expansion. The use of a Legendre expansion removes the dependence on the quadrature from the data. This allows any quadrature to work with any cross section dataset. A $l$-order Legendre polynomial is denoted by $P_l$. Details of the Legendre polynomials can be found in Appendix~\ref{appdx:leg}. The anisotropic scatter cross section is rewritten as a Legendre expansion as
\begin{equation} \label{eq:leg_1}
\Sigma_{s, i, a, a'}^{g, g'} = \sum_{l=0}^L \frac{2l+1}{4 \pi}\Sigma_{s, i, l}^{g, g'} P_l(\mu_s).
\end{equation}

>>>>

%%
%% Edward T. Norris
%% Discrete Ordinates Computed Tomography Organ Dose Simulator (DOCTORS)
%% 
%% === Methods ===
%%

@@

The linear Boltzmann equation (LBE) is generally true of any particle for which outside forces (electromagnetic or gravitational) are either not present or negligible and particle-particle collisions are insignificant. Therefore, the LBE is generally not valid for charged particles or systems whose particle density is on the order of the confining medium. In such systems, particle-particle interactions may not be negligible requiring the non-linear Boltzmann equation.

@@

The steady state form of the LBE can be readily derived from simple intuition. In the steady state, particle production and removal must be equal. In any volume, three production mechanisms are present. Particles can (1) stream into the volume across a bounding surface, (2) inscatter from another energy-direction component of phase space, or (3) be produced directly by the external source. At the same time, three removal terms are present, particles can (1) stream out of the volume of interest by crossing a bounding surface, (2) be absorbed and vanish entirely, or (3) outscatter to a energy-direction component of phase space other than the one of interest.

The downscatter and absorption removal terms are typically grouped together as
\begin{equation}
\Sigma_t = \Sigma_a + \Sigma_s
\end{equation}
where $\Sigma_t$, $\Sigma_a$, and $\Sigma_s$ are the macroscopic total, absorption, and scatter cross sections respectively. The removal of particles from interactions (scatter and absorption) per volume is then
\begin{equation}
\Sigma_t(\boldsymbol{r}, E) \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}
where $\psi$ is the angular flux. The production and removal streaming operators are combined as 
\begin{equation}
\hat{\Omega} \cdot \nabla \psi(\boldsymbol{r}, E, \hat{\Omega})
\end{equation}
and included as a removal term since the normal is defined pointing outward from the volume. The normal sign convention results in particles streaming out being positive and those streaming in becoming negative.

@@

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/coord_sys}
  \end{center}
  \caption{The coordinate system used in DOCTORS. Given an arbitrary direction, $\Omega$, $\mu$, $\eta$, and $\xi$ are its direction cosines with respect to the $x$, $y$, and $z$ axes respectively. $\varphi$ is the azimuthal angle (with respect to $x$) and $\theta$ is the polar angle (with respect to $z$).}
\label{fig:coord_sys}
\end{figure}

The rotation symmetrical ($S_N$) quadrature is implemented in DOCTORS. From amongst many different quadrature sets, the rotation symmetrical quadratures were selected because they are easy to implement, rotationally symmetric, and used in production discrete ordinate solvers~\citep{ref:evanst} which enables simple, direct comparison to other solvers. Arbitrary quadratures are permissible in a plain text file supplied by the user enabling other quadratures.

ADD THEORY AND MATH OF ANGLE/WEIGHT GENERATION?

To include a user defined quadrature, the plain text file should be formatted such that it contains four columns similar to Table~\ref{tab:quad_format} (without any headings). No checks are performed to ensure that octants are covered, this allows simple testing quadratures that utilize only a single direction to be used. Values for $\mu$, $\eta$, and $\xi$ do not necessarily need to add to unity as they will be automatically normalized by DOCTORS. Likewise, the weights will be normalized such that their summation is unity. However, user defined directions \textit{should not} be parallel to any major axis. Doing so may result in undefined behavior. Also, after normalization, no two rows should have identical values for $\mu$, $\eta$, and $\xi$ such as the last two rows of Table~\ref{tab:quad_interp}, nor may any weights be negative. The data in Table~\ref{tab:quad_format} would be read in and renormalized by DOCTORS resulting in data identical to  Table~\ref{tab:quad_interp}.

Values are read in as a 32-bit IEEE-754 floating point number~\citep{ref:ieee754} which ensures seven significant digits. Significant digits beyond this will be truncated. Note that Table~\ref{tab:quad_interp} rounds four digits after the decimal.

@@

A given direction, $\hat{\Omega}$, is determined by its three cosine components: 
\begin{equation} \label{eq:omega_cos}
\hat{\Omega} = \mu \hat{i} + \eta \hat{j} + \xi \hat{k}
\end{equation}
where $\hat{i}$, $\hat{j}$, and $\hat{k}$ are the unit directions in the $x$, $y$, and $z$ directions respectively. Therefore, any discrete direction $\hat{\Omega}_a$ can be expressed as $<\mu_a, \eta_a, \xi_a>$. The two forms will be used interchangeably.

Discretizing the LBE (Eq.~\ref{eq:boltz}) in angular space gives
\begin{equation} \label{eq:boltz_a}
\begin{split}
&\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t(\boldsymbol{r}, E) \right]
\psi_{a}(\boldsymbol{r}, E) = \\
&\sum_{a=0}^{N_a-1} \int_0^\infty \Sigma_{s, a, a'}(\boldsymbol{r}, E' \rightarrow E) \psi_{a'}(\boldsymbol{r}, E') dE' \omega_a + S_a(\boldsymbol{r}, E)
\end{split}
\end{equation}
where the subscript $a$ denotes the direction number and $\omega_a$ is a weight associated with that direction. The weights are designed with the angles such that once a discrete quadrature is selected, integration over continuous space becomes integration in discrete space approximated by
\begin{equation} \label{eq:disc_int}
\int_{4 \pi} f(\hat{\Omega}) d\hat{\Omega} \approx \sum_{a=0}^{N_a-1} f(\hat{\Omega}_a) \omega_a = 1.
\end{equation}

@@

The user must provide cross section data in the form of an AMPX formatted binary file (Format detaisl are provided in Section~\ref{sec:xsparse}). The cross section data file contains group structure information. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight the energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries us a uniform function [CITE].
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

In photon based problems, upscatter is typically negligible [CITE]. By the time photons become low enough in energy to upscatter significantly, they are no longer of dosimetric concern. Removing upscatter reduces Eq.~\ref{eq:boltz_e} to
\begin{equation} \label{eq:boltz_e2}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}
which nearly identical; only the starting point for the scatter summation changes. Computationally, though, the lack of upscatter makes a significant difference. The highest energy group can be solved independently of all those below it. Once that group is solved, the next group can also be solved and so forth. In the presence of upscatter, the highest group depends on the solution to all groups below it so all energy groups must be solved simultaneously.

@@

The problem domain for a CT voxel phantom is a regular, rectangular parallelepiped of dimension $D_x \times D_y \times D_z$, the values of which are known based on the CT setup provided by the user. The mesh is partitioned into $N_x$, $N_y$, and $N_z$ evenly spaced bins along each major direction as shown in Fig.~\ref{fig:spatial_disc}. The total number of voxels, $N_V$ is easily computed from the number of bins in each major direction:
\begin{equation} \label{eq:n_v}
N_V = N_x N_y N_z.
\end{equation}
The length of a voxel along the $x$-direction is computed by 
\begin{equation} \label{eq:mesh_x}
\Delta x = \frac{D_x}{N_x}.
\end{equation}
Analogous equations apply in the $y$ and $z$ directions as well.

@@

Voxels can be indexed in one of two ways, either by their component indices or global index. Each voxel in the mesh has three components indices, one in each major direction, $i_x$, $i_y$, and $i_z$ that range from 0 to $N_x - 1$, $N_y - 1$, or $N_z-1$. However, this makes indexing cumbersome, so they are combined into a unique global index, $i$ using 
\begin{equation} \label{eq:indx_flat}
i = i_x (N_z + N_y) + i_y N_z + i_z.
\end{equation}
In general, this is known as "flattening" a matrix into a one-dimensional vector. More details about the flattening implementation used in DOCTORS is given in Section~\ref{sec:flatten}. Using global indexing discussed, the fully discretized the steady state LBE given is
\begin{equation} \label{eq:boltz_i}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_{t,i}^g \right]
\psi_{i,a}^{g} = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g.
\end{equation}

@@

To solve the fully discretized form of the Linear Boltzmann Equation given in Eq.~\ref{eq:boltz_i}, the gradient operator must first be computed numerically. Recall that $\Omega_a$ can be written in vector notation using its cosine components as defined in Eq.~\ref{eq:omega_cos}. Also recall the definition of the gradient:
\begin{equation} \label{eq:grad}
\nabla f(x, y, z) = \frac{\partial f(x, y, z)}{\partial x} \hat{i} + \frac{\partial f(x, y, z)}{\partial y} \hat{j} + \frac{\partial f(x, y, z)}{\partial z} \hat{k}.
\end{equation}
To a first order approximation, a partial derivative is computed
\begin{equation} \label{eq:deriv_1}
\frac{\partial f(\zeta)}{\partial \zeta} \approx \frac{f(\zeta+\Delta \zeta/2) - f(\zeta + \Delta \zeta/2)}{\Delta \zeta}
\end{equation}
with respect to $\zeta$. Combining Eq.~\ref{eq:grad} and~\ref{eq:deriv_1}, the $\hat{\Omega} \cdot \nabla \psi$ term can be rewritten as:
\begin{equation} \label{eq:spatial_1}
\begin{split}
\Omega \cdot \nabla \psi 
& = <\mu, \eta, \xi> \cdot <\frac{\partial \psi}{\partial x}, \frac{\partial \psi}{\partial y}, \frac{\partial \psi}{\partial z}> \\
& =
\frac{\partial \psi}{\partial x}\mu + \frac{\partial \psi}{\partial y}\eta + \frac{\partial \psi}{\partial z}\xi \\
& \approx 
\frac{\psi(x + \Delta x/2, y, z) - \psi(x - \Delta x/2, y, z)}{\Delta x} \mu \\
&+ 
\frac{\psi(x, y + \Delta y/2, z) - \psi(x, y - \Delta y/2, z)}{\Delta y} \eta \\
&+ 
\frac{\psi(x, y, z + \Delta z/2) - \psi(x, y, z - \Delta z/2)}{\Delta z} \xi.
\end{split}
\end{equation}
The flux variables $\psi_{i,a,x,in}^g$, $\psi_{i,a,x,out}^g$, $\psi_{i,a,y,in}^g$, $\psi_{i,a,y,out}^g$, $\psi_{i,a,z,in}^g$, and $\psi_{i,a,z,out}^g$ are \textit{surface} averaged flux values while $\psi_{i,a}^{g}$ is a \textit{volume} averaged flux value.

The gradient is computed with respect to the direction $\hat{\Omega}$, thus Eq.~\ref{eq:spatial_1} is valid only for the first octant ($\mu, \eta, \xi > 0$). In other octants, a new gradient must be used. Rather than enumerating all eight variants, Eq.~\ref{eq:spatial_1} can be generalized to
\begin{equation} \label{eq:deriv_2}
\Omega \cdot \nabla \psi \approx 
\frac{\psi_{x,out} - \psi_{x,in}}{\Delta x} \mu + 
\frac{\psi_{y,out} - \psi_{y,in}}{\Delta y} \eta + 
\frac{\psi_{z,out} - \psi_{z,in}}{\Delta z} \xi
\end{equation}
where
\begin{equation}
\psi_{x,out} = 
\begin{cases}
\psi(x+\Delta x/2, y, z) \,, \quad \mu > 0 \\
\psi(x-\Delta x/2, y, z) \,, \quad \mu < 0
\end{cases}
\end{equation}
and
\begin{equation}
\psi_{x,in} = 
\begin{cases}
\psi(x-\Delta x/2, y, z) \,, \quad \mu > 0 \\
\psi(x+\Delta x/2, y, z) \,, \quad \mu < 0
\end{cases}
\end{equation}
and $\psi_{y,out}$, $\psi_{y,in}$, $\psi_{z,out}$, and $\psi_{z,in}$ are similarly defined but with respect to $\eta$ and $\xi$ respectively.

@@

The diamond difference approximation, relates the incoming and outgoing surface fluxes to the cell averaged flux. The cell averaged flux is assumed to be the average of any two opposite surface fluxes. This is assumed to be the case in all three spatial dimensions leading to:
\begin{equation} \label{eq:dd}
\begin{split}
\frac{\psi_{i,a,x,in}^g + \psi_{i,a,x,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,y,in}^g + \psi_{i,a,y,out}^g}{2} &= \psi_{i,a}^{g} \\
\frac{\psi_{i,a,z,in}^g + \psi_{i,a,z,out}^g}{2} &= \psi_{i,a}^{g}.
\end{split}
\end{equation}
Multiplying both sides of Eq.~\ref{eq:dd} by 2 and subtracting twice the incoming surface flux from each yields:
\begin{equation} \label{eq:dd2}
\begin{split}
\psi_{i,a,x,out}^g - \psi_{i,a,x,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,x,in}^g \\
\psi_{i,a,y,out}^g - \psi_{i,a,y,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,y,in}^g \\
\psi_{i,a,z,out}^g - \psi_{i,a,z,in}^g &= 2\psi_{i,a}^{g} - 2\psi_{i,a,z,in}^g.
\end{split}
\end{equation}

The form of the left hand side of Eq.~\ref{eq:dd2} is the same as in Eq.~\ref{eq:boltz_i2}. This allows Eq.~\ref{eq:dd2} to be used to make a substitution in Eq.~\ref{eq:boltz_i2} that removes the dependence on the outgoing flux. After the substitution:
\begin{equation} \label{eq:boltz_i3}
\begin{split}
&\left[ 
\frac{2\psi_{i,a}^{g} - 2\psi_{i,a,x,in}^g}{\Delta x} \mu + 
\frac{2\psi_{i,a}^{g} - 2\psi_{i,a,y,in}^g}{\Delta y} \eta + 
\frac{2\psi_{i,a}^{g} - 2\psi_{i,a,z,in}^g}{\Delta z} \xi
\right]
+ \Sigma_{t,i}^g \psi_{i,a}^{g} \\
& = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g.
\end{split}
\end{equation}

@@

\subsection{Convergence Criteria}
The typical convergence criteria used to determine whether or not to terminate further iteration is the maximum relative error from the previous iteration to the current iteration given in Eq.~\ref{eq:conv} over the phase space $\mathbb{R}$ where $t$ is the current iteration number. Once the relative error in the parameter of interest, $\phi$, is below some threshold, $\epsilon$, the iteration stops and the solution is considered converged.

\begin{equation}\label{eq:conv}
\max_{\mathbb{R}} \left\{ \frac{\phi_t - \phi_{t-1}}{\phi_t} \right\} < \epsilon
\end{equation}

In some problems, the convergence criteria given in Eq.~\ref{eq:conv} is not reached for many iterations. In these cases, terminating the solution early rather than waiting for full convergence is appropriate. This is enforced by a second criterial given by Eq.~\ref{eq:conv2}. Whenever the number of iterations, $t$, exceeds the total permissible number of iterations, $T$, the iteration is terminated.

\begin{equation}\label{eq:conv2}
t > T
\end{equation}

@@

ADD FIGURE TO SHOW CONVERGENCE RATE.

In order to measure the uncertainty due to floating-point error, another convergence criteria was added. Rather than considering the relative change from iteration to iteration, the total absolute change is measured:
\begin{equation}\label{eq:conv3}
\varepsilon = \sum_{\mathbb{R}} |\phi_i - \phi_{i-1}|.
\end{equation}
The termination criteria is given by Eq.~\ref{eq:conv4}. Computation of a value for $\varepsilon$ requires two previous iteration, thus this condition cannot be met until the third iteration and thereafter.

\begin{equation}\label{eq:conv4}
\varepsilon_i > \varepsilon_{i-1}
\end{equation}

@@

The data files containing the scatter cross sections stor the information using Legendre polynomial expansion coefficients. The use of a Legendre expansion removes the dependence on the quadrature from the data. This allows any quadrature to work with any cross section dataset. A $l$-order Legendre polynomial is denoted by $P_l$. Details of the Legendre polynomials can be found in Appendix~\ref{appdx:leg}. The anisotropic scatter cross section is rewritten using a Legendre expansion as
\begin{equation} \label{eq:leg_1}
\Sigma_{s, i, a, a'}^{g, g'} = \sum_{l=0}^L \frac{2l+1}{4 \pi}\Sigma_{s, i, l}^{g, g'} P_l(\mu_s).
\end{equation}
